### 2025-05-28 chapters/background.tex
updated relevant chapters & sections

\section{Trusted Execution Environments}

@@

>>>>

\section{Trusted Execution Environments}
Trusted Execution Environments (TEEs) are isolated computing environments that provide strong security guarantees for code and data execution, even in the presence of potentially compromised operating systems and applications. They achieve this by leveraging hardware-based isolation mechanisms to establish secure boundaries, often called enclaves, where sensitive computation can occur without interference from the rest of the system. TEEs aim to ensure confidentiality and integrity through careful hardware and software co-design, minimizing the trusted computing base (TCB) and mitigating a wide range of software attacks. While several commercial TEEs, such as Intel Software Guard Extensions (SGX) and ARM TrustZone, demonstrate the feasibility and practical relevance of enclave-based secure execution, their closed-source and inflexible designs limit research and hardware/software co-design opportunities. This has motivated the development of open TEE frameworks like Keystone, built on RISC-V—an open and extensible instruction set architecture. A key architectural feature of TEEs is privilege separation, where a minimal TCB executes in a highly privileged mode controlling access to critical physical resources, while the operating system and applications run in less privileged, untrusted modes.

@@

Keystone represents an open-source framework designed to facilitate the architecture and implementation of Trusted Execution Environments (TEEs) leveraging hardware enclaves on RISC-V processors. Its primary objective is to enable the construction of customizable TEEs optimized for specific RISC-V platforms, thereby achieving enhanced performance, reduced trusted computing base (TCB) complexity, and improved programmability tailored to distinct workloads and security threat models.

A typical Keystone-capable system architecture comprises multiple components distributed across different privilege modes of the RISC-V privilege hierarchy. The foundational hardware is a trusted CPU package integrating Keystone-compatible RISC-V cores and a silicon root of trust, potentially augmented by features such as cache partitioning, memory encryption, and secure randomness sources. The core security functionality is governed by the Security Monitor (SM), a minimalistic M-mode software component embodying the system's TCB. The SM is responsible for enclave lifecycle management and enforces isolation boundaries between enclaves and the untrusted operating system (OS).

Keystone’s Security Monitor (SM), executing at the highest privilege level (M-mode), serves as the linchpin of the system’s trusted computing base. Beyond enclave lifecycle management, the SM enforces fine-grained memory isolation policies and orchestrates enclave scheduling, maintaining strict control over resource access to prevent unauthorized interference.

\begin{figure}[htbp]
    \centering
    \includegraphics[width=0.9\linewidth]{figures/keystone_overview.png}
    \caption{Overview of the Keystone architecture illustrating components such as the Security Monitor, enclave runtime, and the privilege hierarchy.}
    \label{fig:keystone_overview}
\end{figure}

Enclaves, the fundamental isolation units in Keystone, operate within dedicated physical memory regions inaccessible to the OS and other enclaves. Physical Memory Protection (PMP), a hardware-assisted mechanism controlled by the SM, restricts access to enclave memory exclusively to the enclave and the SM, thus preserving confidentiality and integrity even in the event of OS compromise.

Each enclave comprises two principal layers: the user-level enclave application (eapp) and a supervisor-level runtime environment. The eapp executes user-specific logic within the enclave, while the runtime, operating in supervisor mode (S-mode), manages system calls, exception handling, and virtual memory services intrinsic to enclave operation. This layered design provides a clear separation of concerns, reducing attack surfaces and allowing for customized security policies adapted to workload requirements.

Keystone’s workflow delineates distinct roles for platform providers and enclave developers, fostering modularity and flexibility. Platform providers undertake the compilation and deployment of the SM tailored to the target hardware, ensuring integration of the root of trust and hardware-specific functionalities. Enclave developers leverage the Keystone SDK to construct enclave applications alongside their runtimes and host binaries, which are subsequently packaged and deployed on the target platform independently of underlying hardware specifics.

Furthermore, Keystone supports remote attestation mechanisms, enabling verification of enclave authenticity and integrity prior to provisioning sensitive data or executing critical workloads. This capability is essential for secure deployment in distributed and cloud environments.

The enclave lifecycle progresses through three stages: creation, execution, and destruction. Creation involves the host allocating enclave private memory (EPM), populating it with enclave page tables, runtime, and application binaries, and invoking the SM to isolate the enclave via PMP. During execution, the SM orchestrates transitions into and out of the enclave, dynamically adjusting PMP permissions to maintain strict isolation. Destruction securely clears the enclave’s EPM and reclaims resources, ensuring no residual data remains.

\begin{figure}[htbp]
    \centering
    \includegraphics[width=0.9\linewidth]{figures/enclave_lifecycle.png}
    \caption{Stages of a Keystone enclave lifecycle: creation, execution, and destruction.}
    \label{fig:enclave_lifecycle}
\end{figure}

Keystone’s design also facilitates extensibility, allowing for integration of advanced security features such as secure I/O, cryptographic accelerators, and hardware-assisted debugging. This adaptability future-proofs the architecture, enabling continuous evolution of TEE capabilities aligned with emerging threats and diverse application demands on RISC-V platforms.

### 2025-05-28 chapters/methodology.tex
updated relevant chapters & sections

\section{Experimental setup}

\section{Benchmarking tools and metrics}
%\section{Performance metrics}

\section{Parameter Variation Strategy}

>>>>

\section{Experimental setup}
%The experimental setup utilizes the Keystone-enhanced QEMU emulator running on an Ubuntu host operating system. This configuration allows for flexible development and testing of Keystone enclaves in a controlled environment without requiring physical RISC-V hardware. The Ubuntu host provides a stable platform for building and deploying enclave binaries, and it facilitates the integration of the Keystone kernel driver and SDK.

The Keystone framework was evaluated using a RISC-V specific fork of the QEMU emulator on an Ubuntu 22.04 host system. This configuration allowed testing of Keystone enclaves in a controlled virtualized environment, avoiding the need for physical RISC-V hardware. The system was configured to emulate the RV64 architecture for evaluating Keystone’s performance.

The build process for Keystone utilized the Make and Buildroot tools. Key configuration options included the selection of the QEMU virtual platform, the RV64 architecture, and the corresponding Buildroot configuration. Keystone was initiated in the QEMU environment with the following command:

\begin{verbatim}
make run
\end{verbatim}

This command launched the Security Monitor and the Linux OS. Within the QEMU guest, the Keystone kernel driver was manually loaded using:

\begin{verbatim}
modprobe keystone-driver
\end{verbatim}

For debugging, Keystone was run in debug mode with the QEMU GDB server connected to inspect memory protection and control registers:

\begin{verbatim}
KEYSTONE_DEBUG=y make run
make debug-connect
\end{verbatim}

Additional details on the build process, dependencies, and configuration options are available in the \textbf{Appendix}.

\section{Benchmarking tools and metrics}
%\section{Performance metrics}
The performance characterization of Keystone utilized two established synthetic benchmarks tailored for embedded system evaluation: \textit{Dhrystone} and \textit{CoreMark}. Both were selected for their cross-platform compatibility and established use in RISC-V environments.

\textbf{Dhrystone} evaluates integer arithmetic and control flow performance, making it effective for assessing core CPU throughput in environments where floating-point operations are secondary. The RISC-V version, sourced from the \texttt{riscv-tests} repository, was employed to calculate \textit{Dhrystones per second (DMIPS)}. To isolate Keystone’s overhead, measurements were conducted in both native and enclave contexts.

\textbf{CoreMark}, developed by EEMBC, provides a broader performance assessment through representative workloads, including matrix operations, list processing, and state machine evaluations. The RISC-V implementation from the SiFive repository was used to determine \textit{iterations per second (IPS)}. This benchmark offers a deeper view into system performance under more diverse computational tasks than Dhrystone.

Performance variability was assessed by calculating the standard deviation across repeated runs. Metrics collected included:
\begin{itemize}
\item Execution Time
\item DMIPS (Dhrystone) and IPS (CoreMark)
\item Standard Deviation across trials
\end{itemize}

This dual-benchmark approach ensured a comprehensive evaluation of both low-level CPU operations and higher-level application performance within the enclave context.
\section{Parameter Variation Strategy}

### 2025-05-28 main.tex
Updated README.md to reflect changes in LaTeX structure and preamble.tex
updated relevant chapters & sections

\documentclass[english, version-2020-11]{uzl-thesis}
\UzLStyle{computer modern oldschool design}
\input{preamble.tex} % Put your \UzLStyle and \UzLThesisSetup here

@@

\chapter{Introduction}
\input{chapters/01_introduction.tex}

\chapter{Background and System Architecture}
\input{chapters/02_background.tex}

\chapter{Methodology}
\input{chapters/03_methodology.tex}

\chapter{Results and discussion}
\input{chapters/04_results.tex}

\chapter{System Configuration Recommendations}
\input{chapters/05_recommendations.tex}

\chapter{Conclusion and Future Work}
\input{chapters/06_conclusion.tex}

>>>>

\documentclass[english, version-2020-11]{uzl-thesis}
\UzLStyle{computer modern oldschool design}
\usepackage{float}

\input{preamble.tex} % Put your \UzLStyle and \UzLThesisSetup here

@@

\chapter{Introduction}
\input{chapters/introduction.tex}

\chapter{Background and System Architecture}
The RISC-V instruction set architecture (ISA) serves as a promising foundation for TEE implementation due to its open and extensible nature. RISC-V offers unique benefits beyond its open-source status: it incorporates security-oriented primitives such as Physical Memory Protection (PMP) which facilitates efficient and fine-grained memory isolation. Moreover, its community-driven evolution fosters innovation in security features, enabling frameworks like Keystone to experiment with, and contribute to, ISA enhancements.

RISC-V defines three primary privilege modes: user (U-mode), supervisor (S-mode), and machine (M-mode). These modes govern the execution privileges and resource accessibility of software components. M-mode, the highest privilege level, has exclusive control over hardware resources and interrupt management, rendering it an ideal candidate for hosting the TCB, such as the Security Monitor (SM) in Keystone. Unlike hardware microcode, M-mode software can be developed and patched flexibly using standard programming languages and toolchains, improving maintainability and formal verification potential.

A key hardware feature leveraged by Keystone is the PMP, introduced in the RISC-V Privilege Specification version 1.10. PMP enables M-mode to enforce physical memory access restrictions on lower privilege levels, thus underpinning enclave memory isolation. Despite variations in PMP implementations across platforms, its guarantees are standardized, facilitating portable and secure enclave deployment.

Interrupt and exception handling in RISC-V is primarily managed at the M-mode level, which may delegate common traps, such as page faults and system calls, to S-mode to optimize performance. Virtual memory support is provided by a memory management unit (MMU) that translates virtual addresses to physical addresses using multi-level page tables. While the operating system controls virtual memory via the \texttt{satp} register, Keystone enclaves maintain protected page tables and are secured against OS-based attacks through PMP-enforced physical memory restrictions.

Collectively, these architectural features establish a robust foundation for implementing secure and efficient TEEs on RISC-V platforms, as exemplified by Keystone.

\input{chapters/background.tex}

\chapter{Methodology}
The objective of the performance study was to quantify the overhead introduced by Keystone's enclave isolation mechanisms on RISC-V architectures. Experiments were executed within a QEMU-based virtual environment configured for the RV64 architecture, providing consistency and hardware independence.

To run benchmarks, a separate enclave application (eapp) and corresponding host application were developed for each benchmark. The host application initialized and managed the enclave using the Keystone C++ API, while the enclave application contained the benchmark logic and communicated with the host through edge calls. This modular structure allowed each benchmark to be executed both natively (outside the enclave) and securely (within the enclave), enabling accurate measurement of the enclave-related overhead.
To achieve a comparative analysis, two execution modes were considered:
\begin{enumerate}
\item \textit{Native Execution}: Benchmark execution without enclave isolation.
\item \textit{Enclave Execution}: Benchmark execution within the Keystone TEE.
\end{enumerate}

Each benchmark—Dhrystone and CoreMark—was run multiple times under both conditions. Performance metrics, including total execution time, throughput (DMIPS or IPS), and result variability (standard deviation), were recorded and analyzed.

The evaluation focused on determining the relative performance degradation attributable to enclave isolation by comparing results across the two execution modes.

\input{chapters/methodology.tex}

\chapter{Results and discussion}
\input{chapters/results.tex}

\chapter{System Configuration Recommendations}
\input{chapters/recommendations.tex}

\chapter{Conclusion and Future Work}
\input{chapters/conclusion.tex}
